Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-29 manuscript.rmd
Added tables in chunks

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom, kableExtra))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012]

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{TLP}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

@@

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{TLP}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; [Table 1](#param)]. We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve scheme, we used parameters of a soil water retention curve that we derived from existing data on volumetric water content \theta vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), \theta_{s} is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib; daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (n = 5, cylinder volume = 1250 cm{3}). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES. This figure will be reformatted to compose individual figures into one.")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{TLP}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{TLP}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems. High Kmax is associated with photosynthetic rates with faster growth potential.

@@

## Acknowledgements
## Data Availability

#### Table 1: ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale. No. 1-6 are parameters specific to FATES.  {#param}

| No. | Parameter | Description | Range & Units | Method of determination with reference |
| -- | ------ | ------------- | ------------- |  ------------------------------------------------------------------------------------------------- | 
| 1 | fates_leaf_ BB_slope | stomatal slope parameter, as per Ball-Berry | 4 - 16 Unitless  | The range of Ball Berry parameter fitted across several studies as compiled in Table 2 of [@Medlyn:2012] | 
| 2 | fates_leaf_ slatop | Specific Leaf Area (SLA) at the top of canopy, projected area basis | 0.0042 - 0.0400 $m^2gC^{-1}$ | Based on the observed range of Leaf Mass Area (LMA)--24.97 to 235.8 $gm^{-2}$-- for individual tree variation across 51 tree species in Barro Colorado Island (this study), and assumption of 50% carbon content in biomass. | 
| 3 | fates_leaf_ vcmax25top | Maximum carboxylation rate of Rubisco at $25^\circ$C, canopy top | 22.7 - 92.5 umol $CO^2m^2s^{-1}$ | Observed range of $V_{cmax}$ at  $25^\circ$C for tropical tree species under relative radiation of 50% or greater, thereby excluding highly shaded leaves. Values derived with a conversion specific for CLM model are used. [@Ali:2015] | 
| 4 | fates_roota_ par | CLM rooting distribution parameter | 5.9 - 7.4 per m | Parameter a in Eq. 2 in [@Zeng:2001] that regulates the shape of the rooting profile. Range corresponds to this parameter specified for BATS (or IGBP) land cover classified Deciduous Broadleaf Trees (5.9 $m^{-1}$ and Evergreen Broadleaf Trees  (7.4 $m^{-1}$) as given in Table 1 of [@Zeng:2001]. | 
| 5 | fates_rootb_ par | CLM rooting distribution parameter | 0.217 - 1.956 $m^{-1}$ | Parameter b in Eq. 2 in [@Zeng:2001] that regulates the depth of the rooting profile. Chosen range of *b* is derived using this equation so as to fit the observed range of rooting depth (d~r~) of 2 - 18 m for Tropical Deciduous Forest (mean ± Standard Error (SE); 3.7 ± 0.5, *n* = 5 trees; min = 2, max = 4.7)  and Tropical Evergreen Forest (mean $\pm$ S.E.; 7.3 $\pm$ 0.5, n = 3 trees and 3 communities; min = 2, max = 18 m) combined [@Canadell:1996]. Besides the direct observation of roots at 18 m included by [Nepstad:1994] in Paragominas, eastern Amazonia that is included in the above study; in Tapajos, eastern Amazonia water extraction by roots was also inferred up to 18 m. [@Davidson:2011] | 
| 6 | fates_smpsc | Soil water potential at full stomatal closure | 113000 - 242000 mm | Based on observed range of -1.13 to -2.42 MPa for leaf turgor loss point for individual tree variation across 49 BCI tree species. This study.| 
| 7 | aveDTB | Distance to bedrock | 3 - 18 m | Ben Turner; pers. comm. | 
| 8 | fmax | The maximum fractional saturated area |  unitless | Empirical | 
| 9 | HKSAT_ADJ | Adjusting factor for soil hydraulic conductivity | 1 - 8 unitless | To account for high macroporosity and direct flow paths in tropical soils that is not accounted for by small soil core samples [@Broedel:2017; @Kinner:2004; @Tomasella:1998] | 
| 10 | HKSAT | Soil hydraulic conductivity profile | 0.007 - 0.014 mm/s for =< 12.5 cm; 5.56e-05, 0.0003 for >= 60 cm | For Conrad catchment observed median Ksat (95% CI) for 12.5 cm depth is 38.3 mm/hr (25.4 - 51.2, n = 75), while for 60 cm depth 0.7 mm/hr (0.2 - 1.2, n = 40) [@Kinner:2004]. For reference, a storm with 12.5 mm/hr rainfall intensity has a 0.2 probability of occuring in any given rainfall event. | 
| 11 | fpi_max | Maximum interception fraction of precipitation | 0.05 - 0.44 unitless | Based on throughfall data from [@Zimmermann:2010kzl] and precipitation data for BCI from STRI Physical Monitoring program, defined for precipitation events greater than 10 mm. |

## Supplementary Information

@@

#### Table S1: Scientific names and codes for tree species for which Effective Rooting Depths were estimated. {#erd-sp}

```{r, results = 'asis', echo = FALSE}
writeLines("td, th { padding : 10px } th { background-color : brown ; color : white; border : 1px solid white; } td { color : brown ; border : 1px solid brown }", con = "mystyle.css")
knitr::kable(erd.sp.names, format = "html")
```

# References

>>>>

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
# load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see @Johnson:2018fad;@Kumagai:2011;@Ivanov:2012)

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

@@

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib; daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (*n* = 5, cylinder volume = 1250 cm{3}). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \ref(tab:param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red lines) versus simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES for daily Volumetric Water Content (VWC) by depth for horizontal probes at three depths (10, 40 and 100 cm; A) and a vertical probe (0-15 cm; B); monthly stream discharge (C) and monthly Evapotranspiration from eddy flux tower (D). Points (red) in (A) show average plot-wide observations of VWC with 95% CI (error bars).[Goodness-of-fit values to be added on the panels.]")}
knitr::include_graphics(file.path("figures/hydro-calib2.jpeg"))
```

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~Stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$).

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~Stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

@@

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: Kmax units as in Fig. 2. Water Potentials are in Mpa.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

```{r erd-mort-changed, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~Leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~Stem}$) in two angiosperm systems. High Kmax is associated with photosynthetic rates with faster growth potential.

@@

## Acknowledgements
## Data Availability
## Supplementary Information

@@

```{r symbols, results = 'asis', echo = FALSE}
knitr::kable(symbols.table, caption = 'The symbol, definition of key hydraulic traits or variables used', booktabs = TRUE, escape = FALSE)
```

```{r param, results = 'asis', echo = FALSE}

knitr::kable(param.table, caption = 'ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale. No. 1-6 are parameters specific to FATES.', booktabs = TRUE, escape = FALSE)
```

```{r spcodes, results = 'asis', echo = FALSE}
knitr::kable(erd.sp.names, caption = 'Scientific names and codes for tree species for which Effective Rooting Depths were estimated.', booktabs = TRUE
)
```
# References
